Subsidiary Recursion in Coq

### 2022-01-12 paper/paper.tex
revising text, description of Mu

An important technical novelty of our approach is a derivation of a
weakened form of positive-recursive type in Coq.  Coq (Agda, and Lean)
restrict datatypes $D$ to be strictly positive: in the type for any
constructor of $D$, $D$ cannot occur to the left of any arrows.  Our
derivation needs to use positive-recursive types, where $D$ may occur
to the left only of an even number of arrows.  The restriction to
strict positivity is enforced because full positive-recursive types
are incompatible with Coq's type theory~\cite{coquand88}.  But we
present a way to derive a weakened form that is sufficient for our
examples (Section~\ref{sec:mu}).  The weakening is to require only
that $F\ \mu$ is a retract of $\mu$, where $\mu$ is the recursive type
and $F\ \mu$ its one-step unfolding.  Usually these types are
isomorphic.  Hence, we dub these \textbf{retractive-positive}
recursive types.  This weakening does have some negative consequences,
but we will see how to handle them.  Our definition of
retractive-positive recursive types makes essential use of impredicate
quantification, and hence cannot be soundly recapitulated in a
predicative theory like Agda's.

@@

Our approach is within a long line of work using ideas from universal
algebra and category theory to describe inductive datatypes and their
recursion principles.  On this approach, one describes transformations
to be performed on data as \emph{algebras}, which can then be
\textit{folded} over data.  The simplest form of algebras, namely
$F$-algebras, are morphisms from $F\ A$ to $A$, for carrier object
$A$.  From a programming perspective, an $F$-algebra is presented
with $F\ A$, and must compute a value of type $A$.

Algebras for our subsidiary recursion are quite a bit more complex than
this.  First, for reasons we will explain further below, the carrier
of the algebra will be an \verb|X : Set -> Set|.  Second, instead of
being given just $F\ A$ as input, the algebra is presented with:

\begin{itemize}
\item a type \verb|R : Set|
\item a function \verb|reveal : R -> C|, which reveals values of type \verb|R| as really having \emph{anchor type} \verb|C|.  This type \verb|C| will either by the abstract type of some outer recursion, or else our development's version of the actual datatype one is recursing over.
\item a function \verb|fold : FoldT Alg R|, which allows one to initiate subsidiary recursions in which the anchor type is \verb|R|.  We will present its type \verb|FoldT Alg R| below.
\item a function \verb|eval : R -> X R|, to use for making recursive calls, on any value of type \verb|R|.
\item and an \emph{subdata structure} \verb|d : F R|, where \verb|F| is the signature functor for the datatype.
\end{itemize}

\noindent The algebra is then required to produce a value of type \verb|F R|.

We use Coq inductive types for the signature functors \verb|F| of
various datatypes, thus enabling recursions to use Coq's
pattern-matching on the subdata structure \verb|d|.  So the style of
coding against this interface retains a similar feel to structural
recursions.  Unlike with structural termination, though, the interface
here is type-based and hence compositional.  As we will see, it
supports nested and higher-order recursions.

As in previous work, we dub this interface a \emph{recursion
universe}~\cite{stump20}.  As in other domains using the term
``universe'', we have an entity (here, \verb|R|) from which one cannot
escape by using the available operations (for other cases:
$\epsilon_0$ and $\omega^-$, the physical universe and traveling at
the speed of light).  Staying in the recursion universe is good,
because we may recurse (via \verb|eval|) on any value of type
\verb|R|.

Some points must still be explained, particularly why \verb|X| has type \verb|Set -> Set|,
and the definition of \verb|FoldT|.  Let us see these details next.

@@

Let us consider two files from our development.

@@

  This file is parametrized by a signature functor \verb|F| of type
  \verb|Set -> Set|.  It provides the implementation of subsidiary
  recursion.  Two crucial values are \verb|Subrec : Set|, which is the
  type to use for subsidiary recursion; and
  \verb|inn : F Subrec -> Subrec|, which is to be used as a
  constructor for that type.  An important point, however, is that
  \verb|Subrec.v| does not provide an induction principle based on
  \verb|inn|.  Induction is derived later
  (Section~\ref{sec:interfacei}). \verb|Subrec.v| makes critical use
  of retractive-positive recursive types, to take a fixed-point based
  on \verb|F|.  We will present the definition of those in
  Section~\ref{sec:mu}.

@@

  This file specializes the development in \verb|Subrec.v| to the case
  of lists (parametrized by the type \verb|A| of elements).  In
  general, to use our development to get subsidiary recursion over
  some datatype, one will have a similar ``shim'' file.  The file
  defines the signature functor \verb|ListF|, shown in
  Figure~\ref{fig:listf}.  Using \verb|Subrec|, we then get a type
  \verb|List|.  This is not to be confused with the type \verb|list|
  of lists in Coq's standard library.  As noted previously, our
  development is meant to be used in extension of existing inductive
  datatypes, not replacing them.  The figure also shows constructors
  \verb|mkNil| and \verb|mkCons| for \verb|List|, and types for
  conversion functions between \verb|List| and \verb|list| (see
  Section~\ref{sec:deriv} for code).

@@

\verb|Subrec.v| also defines the notion of algebra that is used for writing
recursions.  The central definitions are in Figure~\ref{fig:algf}.
\verb|KAlg| is the kind for the type-constructor for algebras, as we
see in the very last definition of the figure, for \verb|Alg|.  This
type-constructor \verb|Alg| is a fixed-point of the type \verb|AlgF|.
The fixed-point is taken using \verb|MuAlg| (Section~\ref{sec:mu}
below).  Doing so requires that \verb|AlgF| only use its parameter
\verb|Alg| positively.  We will confirm this shortly.

@@

Definition Alg : KAlg := MuAlg AlgF.
\end{verbatim}
\caption{The type for algebras (\texttt{Subrec.v})}
\label{fig:algf}
\end{figure}

@@

Let us look at \verb|AlgF|.  As noted already, each recursion is based
on an abstract type \verb|R|, representing the data upon which we will
recurse.  This is the first argument to a value of type
\verb|AlgF Alg C X|.  An algebra can assume nothing about \verb|R|
except that it supports the following operations.  First there is
\verb|reveal|, which turns an \verb|R| into a \verb|C|.  This reveals
that the data of type \verb|R| are really values of the datatype
(\verb|List|, for example), if this is a top-level recursion; or else
really belong to some outer recursion universe, if this is an inner
recursion.  Next we have \texttt{fold}, which will allow us to fold
another algebra over data of type \verb|R|.  We will use \verb|fold|
to initiate subsidiary recursions.  Then there is \verb|eval|, which
is used to make recursive calls on data of type \verb|R|.

@@

@@

@@

Let us consider the code in Figure~\ref{fig:mu}.  The critical
idea is embodied in the definition of \verb|Mu|.  We would like
to have a definition like
\begin{verbatim}
  Inductive Mu' : Set := mu' : F Mu' -> Mu'.
\end{verbatim}
\noindent This is exactly what is used in approaches to modular
datatypes in functional programming, like
Swierstra's~\cite{swierstra08}.  But this definition is (rightly)
rejected by Coq, as it would unsoundly enable instantiations of
\verb|F| that are not strictly positive.

>>>>

An important technical novelty of our approach is a derivation of a
weakened form of positive-recursive type in Coq.  Coq (Agda, and Lean)
restrict datatypes $D$ to be strictly positive: in the type for any
constructor of $D$, $D$ cannot occur to the left of any arrows.  Our
derivation needs to use positive-recursive types, where $D$ may occur
to the left of an even number (only) of arrows.  Coq requires strict
positivity because in the presence of other features of Coq's theory,
full positive-recursive types lead to a paradox~\cite{coquand88}.  We
present a way to derive a weakened form of positive-recursive type
that is sufficient for our examples (Section~\ref{sec:mu}).  The
weakening is to require only that $F\ \mu$ is a retract of $\mu$,
where $\mu$ is the recursive type and $F\ \mu$ its one-step unfolding.
Usually these types are isomorphic.  Hence, we dub these
\textbf{retractive-positive} recursive types.  This weakening has the
negative consequence of leading to a form of noncanonicity, but we
will see how to work around this.  Our definition of
retractive-positive recursive types makes essential use of impredicate
quantification, and hence cannot be soundly recapitulated in a
predicative theory like Agda's.

@@

Our approach is within a long line of work using ideas from universal
algebra and category theory to describe inductive datatypes and their
recursion principles.  On this approach, one describes transformations
to be performed on data as \emph{algebras}, which can then be
\textit{folded} over data.  The simplest form of algebras, namely
$F$-algebras, are morphisms from $F\ A$ to $A$, for carrier object
$A$.  From a programming perspective, an $F$-algebra is given input of
type $F\ A$, and must compute a result of type $A$.

Algebras for our subsidiary recursion are more complex.  First, for
reasons we will explain further below, the carrier of the algebra will
be a functor \verb|X : Set -> Set|.  Second, algebras have a specified
\emph{anchor type} \verb|C|, which we can think of as the datatype
\emph{as viewed by a containing recursion} or else, if this is a
top-level recursion, our development's version of the actual datatype
(e.g., \verb|List|).  The algebra is presented with:

\begin{itemize}
\item a type \verb|R : Set|, which will be this recursion's view of the datatype.
\item a function \verb|reveal : R -> C|, which reveals values of type \verb|R| as really having the anchor type.  
\item a function \verb|fold : FoldT Alg R|, which allows one to initiate subsidiary recursions in which the anchor type is \verb|R|.  Note that the algebra's anchor type is \verb|C|, but for subsidiary recursions the anchor type changes (to \verb|R|). We will present the type \verb|FoldT Alg R| below.
\item a function \verb|eval : R -> X R|, to use for making recursive calls, on any value of type \verb|R|.
\item and a \emph{subdata structure} \verb|d : F R|, where \verb|F| is the signature functor for the datatype.
\end{itemize}

\noindent The algebra is then required to produce a value of type \verb|X R|.

We will use Coq inductive types for the signature functors \verb|F| of
various datatypes, thus enabling recursions to use Coq's
pattern-matching on the subdata structure \verb|d|.  So the style of
coding against this interface retains a similar feel to structural
recursions.  Unlike with structural termination, though, the interface
here is type-based and hence compositional.  As we will see, it
supports nested and higher-order recursions.

As in previous work, we dub this interface a \emph{recursion
universe}~\cite{stump20}.  As in other domains using the term
``universe'', we have an entity (here, \verb|R|) from which one cannot
escape by using the available operations (for other cases:
the ordinal $\epsilon_0$ and $\omega^-$, the physical universe and traveling at
the speed of light).  Staying in the recursion universe is good,
because we may recurse (via \verb|eval|) on any value of type
\verb|R|.

Some points must still be explained, particularly why \verb|X| has type \verb|Set -> Set|,
and the definition of \verb|FoldT|.  Let us see these and other details next.

@@

Let us consider two central files from our development.

@@

  This file is parametrized by a signature functor \verb|F| of type
  \verb|Set -> Set|.  It provides the implementation of subsidiary
  recursion.  Two crucial values are \verb|Subrec : Set|, which is the
  type to use for subsidiary recursion; and
  \verb|inn : F Subrec -> Subrec|, which is to be used as a
  constructor for that type.  An important point, however, is that
  \verb|Subrec.v| does not provide an induction principle based on
  \verb|inn|.  Induction is derived later
  (Section~\ref{sec:interfacei}). \verb|Subrec.v| makes critical use
  of retractive-positive recursive types, to take a fixed-point of a
  construction based on \verb|F|.  We present these recursive
  types in Section~\ref{sec:mu} below.

@@

  This file specializes the development in \verb|Subrec.v| to the case
  of lists (parametrized by the type \verb|A| of elements).  In
  general, to use our development to get subsidiary recursion over
  some datatype, one will have a similar ``shim'' file.  The file
  defines the signature functor \verb|ListF|, shown in
  Figure~\ref{fig:listf}.  Using \verb|Subrec|, we then get a type
  \verb|List|.  This is not to be confused with the type \verb|list|
  of lists in Coq's standard library.  As noted previously, our
  development is meant to be used in extension of existing inductive
  datatypes, not replacing them.  The figure also shows constructors
  \verb|mkNil| and \verb|mkCons| for \verb|List|, and types for
  conversion functions between \verb|List| and \verb|list| (see
  Section~\ref{sec:deriv} for the code).

@@

\verb|Subrec.v| also defines the notion of algebra that is used for
writing recursions.  The central definitions are in
Figure~\ref{fig:algf}.  \verb|KAlg| is the kind for the
type-constructor for algebras, as we see in the definition of
\verb|Alg|.  This type-constructor \verb|Alg| is a fixed-point of the
type \verb|AlgF|.  The fixed-point is taken using \verb|MuAlg|
(Section~\ref{sec:mu}), which implements our retractive-positive
recursive types at kind \verb|KAlg|.  Using \verb|Alg| will require
that \verb|AlgF| only uses its parameter \verb|Alg| positively.  We
will confirm this shortly.

@@

Definition Alg : KAlg := MuAlg AlgF.

Definition fold : FoldT Alg Subrec.
\end{verbatim}
\caption{The type for algebras (\texttt{Subrec.v})}
\label{fig:algf}
\end{figure}

@@

Let us look at \verb|AlgF|.  As noted already, each recursion is based
on an abstract type \verb|R|, representing the data upon which we will
recurse.  This is the first argument to a value of type
\verb|AlgF Alg C X|.  An algebra can assume nothing about \verb|R|
except that it supports the following operations.  First there is
\verb|reveal|, which turns an \verb|R| into a \verb|C|.  This reveals
that the data of type \verb|R| are really values of the anchor type
of this recursion.  Next we have \texttt{fold}, which will allow us to fold
another algebra over data of type \verb|R|.  We will use \verb|fold|
to initiate subsidiary recursions.  Then there is \verb|eval|, for
recursive calls on data of type \verb|R|.

@@

The final definition in the figure is for \verb|fold|, which allows
us to fold an \verb|Alg| over a value of type \verb|Subrec|.  We will
return to the code for this, and definitions of \verb|Subrec| and \verb|inn|,
in Section~\ref{sec:deriv}.

@@

Having seen now the interface for subsidiary recursion in Coq, let us
consider now some examples, listed by their filename in the development.

\subsection{\texttt{Span.v}}

@@

Let us consider the code in Figure~\ref{fig:mu}.  The critical
idea is embodied in the definition of \verb|Mu|.  Ideally, we would like
to have a definition like
\begin{verbatim}
  Inductive Mu' : Set := mu' : F Mu' -> Mu'.
\end{verbatim}
\noindent This is exactly what is used in approaches to modular
datatypes in functional programming, like
Swierstra's~\cite{swierstra08}.  But this definition is (rightly)
rejected by Coq, as instantiations of
\verb|F| that are not strictly positive would be unsound.
